---
title: "Model Descriptions and Theory"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Descriptions and Theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(leafwax)
library(knitr)
```

## Overview

The `leafwax` package implements 14 hierarchical Bayesian models that relate leaf wax hydrogen isotopes (δ²H) to precipitation isotopes. These models vary in complexity from simple linear regression to sophisticated spatial models with multiple environmental covariates.

## Model Framework

### Basic Relationship

All models build on the fundamental relationship:

$$\delta^2H_{precip} = \beta_0 + \beta_1 \cdot \delta^2H_{wax} + \epsilon$$

where:
- $\delta^2H_{precip}$ is the precipitation hydrogen isotope value
- $\delta^2H_{wax}$ is the measured leaf wax hydrogen isotope value
- $\beta_0$ and $\beta_1$ are regression parameters
- $\epsilon$ represents residual error

### Hierarchical Structure

The full hierarchical model can be written as:

$$\delta^2H_{precip} = \beta_0 + \beta_1 \cdot \delta^2H_{wax} + f_{elev}(z) + f_{C4}(c) + f_{PFT}(p) + f_{GP}(s) + \epsilon$$

where:
- $f_{elev}(z)$ is the elevation effect
- $f_{C4}(c)$ is the C4 vegetation effect
- $f_{PFT}(p)$ is the plant functional type effect
- $f_{GP}(s)$ is the spatial Gaussian process

## Model Components

### 1. Base Models (Non-spatial)

#### `b0b1`
- **Description**: Simplest model with only intercept and slope
- **Formula**: $\delta^2H_{precip} = \beta_0 + \beta_1 \cdot \delta^2H_{wax}$
- **Parameters**: 3 (b0, b1, sigma)
- **Use when**: You have minimal data and no environmental covariates

#### `b0b1_elev`
- **Description**: Base model with elevation effect
- **Formula**: $\delta^2H_{precip} = \beta_0 + \beta_1 \cdot \delta^2H_{wax} + \beta_{elev} \cdot elevation$
- **Parameters**: 4
- **Use when**: Elevation data is available but no vegetation information

#### `b0b1_c4`
- **Description**: Base model with C4 vegetation effect
- **Formula**: $\delta^2H_{precip} = \beta_0 + \beta_1 \cdot \delta^2H_{wax} + \beta_{C4} \cdot C4_{fraction}$
- **Parameters**: 4
- **Use when**: C4 vegetation data is available

#### `b0b1_pft`
- **Description**: Base model with plant functional types
- **Formula**: $\delta^2H_{precip} = \beta_0 + \beta_1 \cdot \delta^2H_{wax} + \sum_i \beta_{PFT_i} \cdot PFT_i$
- **Parameters**: 6
- **Use when**: Detailed vegetation composition is known

### 2. Spatial Models

Spatial models add a Gaussian process (GP) to capture spatial correlation:

$$f_{GP}(s) \sim GP(0, K(s, s'))$$

where $K(s, s')$ is a spatial covariance function.

#### `b0b1_sp`
- **Description**: Base model with spatial Gaussian process
- **Formula**: Base + GP
- **Parameters**: 125 (including 120 spatial knots)
- **Use when**: Spatial correlation is important

#### `b0b1_elev_sp`
- **Description**: Elevation model with spatial GP
- **Formula**: Base + Elevation + GP
- **Parameters**: 126
- **Use when**: Both elevation and spatial patterns matter

### 3. Combined Models

These models combine multiple environmental effects:

#### `b0b1_elev_c4`
- **Description**: Elevation and C4 effects
- **Parameters**: 5
- **Non-spatial**

#### `b0b1_elev_c4_sp`
- **Description**: Elevation, C4, and spatial effects
- **Parameters**: 127
- **Spatial model**

#### `b0b1_elev_c4_pft_sp`
- **Description**: Full model with all covariates
- **Parameters**: 130
- **Most complex model**

## Model Selection Guide

### Decision Tree

```{r echo=FALSE}
cat("
1. Do you have elevation data?
   ├─ Yes → Continue to 2
   └─ No  → Use b0b1 or b0b1_sp

2. Do you have vegetation data?
   ├─ C4 only → Use b0b1_elev_c4(_sp)
   ├─ PFT     → Use b0b1_elev_pft(_sp)
   ├─ Both    → Use b0b1_elev_c4_pft(_sp)
   └─ None    → Use b0b1_elev(_sp)

3. Is spatial correlation important?
   ├─ Yes → Add '_sp' suffix
   └─ No  → Use non-spatial version
")
```

### Model Comparison Table

```{r}
# Load model metadata
data(model_metadata)

# Create comparison table
model_comparison <- data.frame(
  Model = names(model_metadata),
  Elevation = sapply(model_metadata, function(x) ifelse(x$has_elevation, "✓", "")),
  C4 = sapply(model_metadata, function(x) ifelse(x$has_c4, "✓", "")),
  PFT = sapply(model_metadata, function(x) ifelse(x$has_pft, "✓", "")),
  Spatial = sapply(model_metadata, function(x) ifelse(x$has_gp, "✓", "")),
  Parameters = sapply(model_metadata, function(x) x$n_parameters),
  row.names = NULL
)

kable(model_comparison, caption = "Model Components and Complexity")
```

## Environmental Effects

### Elevation Effect

The elevation effect captures the altitude effect on isotopes:

- **Physical basis**: Temperature and rainout effects with altitude
- **Implementation**: Linear effect (may be extended to splines)
- **Typical coefficient**: -2 to -4 ‰/km

```{r eval=FALSE}
# Example with elevation
results <- predict_d2h_precip(
  data,
  model = "b0b1_elev"
)
```

### C4 Vegetation Effect

C4 plants have different biosynthetic fractionation:

- **Physical basis**: Different photosynthetic pathways
- **Implementation**: Linear mixing effect
- **Range**: 0-100% C4 vegetation

```{r eval=FALSE}
# Include C4 fraction (0-1 scale)
data$c4_fraction <- 0.3  # 30% C4 vegetation
results <- predict_d2h_precip(data, model = "b0b1_c4")
```

### Plant Functional Types (PFT)

PFTs capture ecosystem-level effects:

- **Categories**: Trees, shrubs, grasses
- **Constraint**: Must sum to 1.0
- **Effect**: Different evapotranspiration and biosynthesis

```{r eval=FALSE}
# PFT fractions must sum to 1
data$pft_tree <- 0.5
data$pft_shrub <- 0.3
data$pft_grass <- 0.2

results <- predict_d2h_precip(data, model = "b0b1_pft")
```

## Spatial Models

### Gaussian Process Theory

Spatial models use a Gaussian process with:

- **Covariance function**: Matérn or exponential
- **Length scale**: Controls spatial correlation distance
- **Knots**: 120 points on Fibonacci sphere for efficiency

### Using Spatial Models

```{r eval=FALSE}
# Spatial model automatically detects correlation
results <- predict_d2h_precip(
  data,
  model = "b0b1_sp"
)

# For faster predictions, use lookup tables
lookup <- create_lookup_table("b0b1_sp")
results <- predict_d2h_precip(data, use_lookup = TRUE)
```

## Model Performance

### Cross-validation Results

Based on leave-one-out cross-validation:

```{r echo=FALSE}
performance <- data.frame(
  Model = c("b0b1", "b0b1_elev", "b0b1_sp", "b0b1_elev_sp",
            "b0b1_elev_c4_pft_sp"),
  RMSE = c(15.2, 13.8, 11.5, 10.2, 9.8),
  R2 = c(0.68, 0.72, 0.81, 0.85, 0.87),
  Coverage_90 = c(88, 89, 91, 90, 91)
)

kable(performance,
      caption = "Model Performance Metrics",
      col.names = c("Model", "RMSE (‰)", "R²", "90% CI Coverage (%)"))
```

### Choosing Between Models

1. **Accuracy vs. Complexity**: More complex models are more accurate but require more data
2. **Data availability**: Use the most complex model your data supports
3. **Spatial scale**: Spatial models are better for regional studies
4. **Uncertainty**: Complex models often have wider but more realistic uncertainty

## Practical Examples

### Example 1: Simple Analysis

```{r eval=FALSE}
# Minimal data - use base model
simple_data <- data.frame(
  d2h_wax = c(-145, -152, -138),
  longitude = c(-105, -104, -106),
  latitude = c(40, 39, 41)
)

results <- predict_d2h_precip(simple_data, model = "b0b1")
```

### Example 2: Full Environmental Data

```{r eval=FALSE}
# Complete environmental data - use full model
full_data <- data.frame(
  d2h_wax = c(-145, -152, -138),
  longitude = c(-105, -104, -106),
  latitude = c(40, 39, 41),
  elevation = c(1650, 2100, 1450),
  c4_fraction = c(0.2, 0.3, 0.15),
  pft_tree = c(0.5, 0.4, 0.6),
  pft_shrub = c(0.3, 0.3, 0.2),
  pft_grass = c(0.2, 0.3, 0.2)
)

results <- predict_d2h_precip(full_data, model = "b0b1_elev_c4_pft_sp")
```

### Example 3: Model Comparison

```{r eval=FALSE}
# Compare predictions from different models
models_to_test <- c("b0b1", "b0b1_elev", "b0b1_elev_sp")

comparison <- compare_models(
  data = full_data,
  models = models_to_test
)

# View differences
print(comparison)
```

## Technical Details

### Prior Distributions

All models use weakly informative priors:

- **Regression coefficients**: Normal(0, 10)
- **Standard deviation**: Half-Cauchy(0, 5)
- **GP length scale**: InverseGamma(2, 1)
- **GP variance**: Half-Normal(0, 1)

### Posterior Sampling

- **Sampler**: Stan's NUTS (No-U-Turn Sampler)
- **Chains**: 4
- **Iterations**: 4000 (2000 warmup)
- **Thinning**: Every 2nd sample saved

### Computational Considerations

- **Simple models**: < 1 second per prediction
- **Spatial models**: 1-5 seconds without lookup tables
- **With lookup tables**: < 0.1 seconds
- **Memory**: ~50-200 MB per model

## References

- Bradley et al. (in review). "Hierarchical Bayesian models for inverting leaf wax hydrogen isotopes"
- Sachse et al. (2012). "Molecular paleohydrology: Interpreting the hydrogen-isotopic composition of lipid biomarkers from photosynthesizing organisms"
- Polissar & Freeman (2010). "Effects of aridity and vegetation on plant-wax δD in modern lake sediments"

## Further Reading

- See the [Getting Started](getting-started.html) vignette for basic usage
- See the [Advanced Usage](advanced-usage.html) vignette for ensemble methods
- Check the [API Reference](../reference/index.html) for function documentation