---
title: "Advanced Usage and Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Usage and Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(leafwax)
```

## Introduction

This vignette covers advanced features of the `leafwax` package including ensemble predictions, custom spatial grids, performance optimization, and integration with other R packages.

## Ensemble Methods

### Why Use Ensembles?

Model ensembles provide more robust predictions by:
- Accounting for structural uncertainty
- Reducing overfitting to specific model assumptions
- Providing more realistic uncertainty estimates

### Basic Ensemble

```{r eval=FALSE}
# Equal-weight ensemble
ensemble_results <- invert_d2H_ensemble(
  data = example_data,
  models = c("b0b1", "b0b1_elev", "b0b1_sp"),
  ensemble_method = "equal"
)

# View ensemble statistics
print(ensemble_results$ensemble_summary)
```

### Weighted Ensemble

Weight models by their cross-validation performance:

```{r eval=FALSE}
# Define model weights based on R²
model_weights <- c(
  b0b1 = 0.68,
  b0b1_elev = 0.72,
  b0b1_sp = 0.81
)

# Normalize weights
model_weights <- model_weights / sum(model_weights)

# Custom weighted ensemble
weighted_ensemble <- function(data, models, weights) {
  all_predictions <- list()

  for (i in seq_along(models)) {
    model_pred <- predict_d2h_precip(data, model = models[i])
    all_predictions[[i]] <- model_pred$d2h_precip_mean * weights[i]
  }

  # Weighted average
  ensemble_mean <- Reduce("+", all_predictions)
  return(ensemble_mean)
}

ensemble_pred <- weighted_ensemble(
  example_data,
  names(model_weights),
  model_weights
)
```

### Bayesian Model Averaging

```{r eval=FALSE}
# Compare models and get ensemble
comparison <- compare_models(
  data = example_data,
  models = c("b0b1", "b0b1_elev", "b0b1_sp", "b0b1_elev_sp"),
  summary_fun = mean,  # or median, weighted.mean
  return_all = FALSE   # Return ensemble summary
)

print(comparison)
```

## Custom Spatial Grids and Lookup Tables

### Creating Regional Lookup Tables

For focused regional studies, create high-resolution lookup tables:

```{r eval=FALSE}
# Western US high-resolution grid
western_us_lookup <- create_regional_lookup(
  model_name = "b0b1_elev_sp",
  lon_range = c(-125, -100),
  lat_range = c(32, 49),
  resolution = 0.25,  # Quarter degree
  n_draws = 500
)

# Use for predictions
results <- get_spatial_params(
  longitude = c(-120, -115, -110),
  latitude = c(45, 42, 38),
  lookup_table = western_us_lookup,
  method = "bilinear"
)
```

### Custom Grid Generation

```{r eval=FALSE}
# Create custom irregular grid
my_grid <- data.frame(
  lon = c(-120, -115, -110, -105, -100),
  lat = c(35, 37, 39, 41, 43),
  cell_id = 1:5
)

# Generate lookup table for custom points
custom_lookup <- create_lookup_table(
  model_name = "b0b1_sp",
  grid = my_grid,
  n_draws = 100
)
```

### Adaptive Grid Resolution

```{r eval=FALSE}
# Variable resolution based on data density
create_adaptive_grid <- function(data, min_res = 0.5, max_res = 2) {
  # Calculate data density
  library(sp)
  library(raster)

  # Create density raster
  coords <- data[, c("longitude", "latitude")]
  sp_points <- SpatialPoints(coords)

  # Determine optimal resolution per region
  # ... implementation details ...

  return(adaptive_grid)
}
```

## Performance Optimization

### Parallel Processing

```{r eval=FALSE}
# Setup parallel backend
library(parallel)
n_cores <- detectCores() - 1

# Parallel batch processing
large_dataset <- data.frame(
  site_id = 1:10000,
  d2h_wax = rnorm(10000, -140, 20),
  longitude = runif(10000, -120, -80),
  latitude = runif(10000, 25, 45)
)

# Process in parallel
results <- batch_predict(
  large_dataset,
  chunk_size = 500,
  parallel = TRUE,
  n_cores = n_cores
)
```

### Memory Management

```{r eval=FALSE}
# Monitor memory during processing
monitor_memory("Before processing")

# Process in memory-efficient chunks
process_large_file <- function(file_path, chunk_rows = 1000) {
  con <- file(file_path, "r")
  header <- readLines(con, n = 1)

  all_results <- list()
  chunk_num <- 1

  while (length(lines <- readLines(con, n = chunk_rows)) > 0) {
    # Process chunk
    chunk_data <- read.csv(textConnection(c(header, lines)))

    # Make predictions
    chunk_results <- predict_d2h_precip(chunk_data)

    # Store results
    all_results[[chunk_num]] <- chunk_results

    # Clean up
    rm(chunk_data, chunk_results)
    gc()

    chunk_num <- chunk_num + 1
    monitor_memory(paste("Chunk", chunk_num))
  }

  close(con)
  return(do.call(rbind, all_results))
}
```

### Caching Strategies

```{r eval=FALSE}
# Setup persistent cache
options(leafwax.cache_dir = "~/leafwax_cache")

# Pre-cache all spatial models
spatial_models <- c("b0b1_sp", "b0b1_elev_sp", "b0b1_c4_sp")

for (model in spatial_models) {
  # Download if needed
  if (!check_data_cache(model, "standard")) {
    download_model_data(model, "standard")
  }

  # Create and cache lookup table
  lookup <- create_lookup_table(model, cache_dir = "~/leafwax_cache")
}

# Now predictions are fast
results <- predict_d2h_precip(data, model = "b0b1_sp", use_lookup = TRUE)
```

## Working with Uncertainty

### Full Posterior Distributions

```{r eval=FALSE}
# Get full posterior draws
full_results <- predict_d2h_precip(
  example_data,
  model = "b0b1_elev",
  return_draws = TRUE,
  n_draws = 2000
)

# Access posterior draws matrix
posterior_draws <- full_results$posterior_draws
dim(posterior_draws)  # n_locations x n_draws

# Custom uncertainty analysis
library(ggplot2)

# Plot full distribution for first site
first_site_draws <- posterior_draws[1, ]

ggplot(data.frame(x = first_site_draws), aes(x = x)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black") +
  geom_vline(xintercept = quantile(first_site_draws, c(0.025, 0.975)),
             linetype = "dashed", color = "red") +
  labs(x = "Precipitation δ²H (‰)",
       y = "Frequency",
       title = "Posterior Distribution - Site 1") +
  theme_minimal()
```

### Propagating Uncertainty

```{r eval=FALSE}
# Propagate measurement uncertainty
propagate_uncertainty <- function(data, n_monte_carlo = 100) {
  n_sites <- nrow(data)
  mc_results <- array(NA, dim = c(n_sites, n_monte_carlo))

  for (i in 1:n_monte_carlo) {
    # Perturb measurements within uncertainty
    perturbed_data <- data
    perturbed_data$d2h_wax <- rnorm(
      n_sites,
      mean = data$d2h_wax,
      sd = data$d2h_wax_sd
    )

    # Get predictions
    pred <- predict_d2h_precip(perturbed_data, verbose = FALSE)
    mc_results[, i] <- pred$d2h_precip_mean
  }

  # Summarize
  total_uncertainty <- data.frame(
    mean = rowMeans(mc_results),
    sd = apply(mc_results, 1, sd),
    lower_95 = apply(mc_results, 1, quantile, 0.025),
    upper_95 = apply(mc_results, 1, quantile, 0.975)
  )

  return(total_uncertainty)
}

uncertainty_results <- propagate_uncertainty(example_data)
```

## Integration with Other Packages

### Integration with sf for Spatial Data

```{r eval=FALSE}
library(sf)

# Convert to spatial object
sites_sf <- st_as_sf(
  example_data,
  coords = c("longitude", "latitude"),
  crs = 4326
)

# Add predictions
predictions <- predict_d2h_precip(example_data)
sites_sf <- cbind(sites_sf, predictions)

# Spatial operations
library(tmap)
tm_shape(sites_sf) +
  tm_dots(col = "d2h_precip_mean",
          size = 0.5,
          palette = "RdBu",
          title = "Precip δ²H") +
  tm_layout(legend.outside = TRUE)
```

### Time Series Analysis

```{r eval=FALSE}
# Analyze temporal trends
library(lubridate)
library(dplyr)

# Add time component
temporal_data <- example_data %>%
  mutate(
    year = rep(2000:2009, each = 1),
    date = ymd(paste(year, "06", "15"))
  )

# Get predictions
temporal_predictions <- predict_d2h_precip(temporal_data)
temporal_data <- cbind(temporal_data, temporal_predictions)

# Analyze trends
library(ggplot2)
ggplot(temporal_data, aes(x = date, y = d2h_precip_mean)) +
  geom_point() +
  geom_smooth(method = "loess") +
  geom_ribbon(aes(ymin = d2h_precip_lower,
                  ymax = d2h_precip_upper),
              alpha = 0.3) +
  labs(x = "Date", y = "Precipitation δ²H (‰)",
       title = "Temporal Reconstruction") +
  theme_minimal()
```

### Machine Learning Integration

```{r eval=FALSE}
# Use leafwax predictions as features
library(randomForest)

# Get predictions from multiple models
features <- data.frame(
  b0b1 = predict_d2h_precip(data, model = "b0b1")$d2h_precip_mean,
  b0b1_elev = predict_d2h_precip(data, model = "b0b1_elev")$d2h_precip_mean,
  b0b1_sp = predict_d2h_precip(data, model = "b0b1_sp")$d2h_precip_mean
)

# Train meta-model
# rf_model <- randomForest(true_values ~ ., data = features)
```

## Custom Workflows

### Workflow 1: Site Selection

```{r eval=FALSE}
# Identify optimal sampling locations
identify_gaps <- function(existing_sites, target_region, min_distance = 50) {
  # Create spatial grid
  grid <- generate_global_grid(
    lon_min = target_region[1],
    lon_max = target_region[2],
    lat_min = target_region[3],
    lat_max = target_region[4],
    resolution = 1
  )

  # Calculate distances to existing sites
  library(geosphere)

  gaps <- grid
  gaps$min_dist <- apply(grid[, c("lon", "lat")], 1, function(pt) {
    dists <- distGeo(pt, existing_sites[, c("longitude", "latitude")])
    min(dists) / 1000  # Convert to km
  })

  # Identify gaps
  gaps <- gaps[gaps$min_dist > min_distance, ]

  return(gaps[order(gaps$min_dist, decreasing = TRUE), ])
}
```

### Workflow 2: Cross-Validation

```{r eval=FALSE}
# Leave-one-out cross-validation
cross_validate_model <- function(data, model_name) {
  n <- nrow(data)
  predictions <- numeric(n)

  for (i in 1:n) {
    # Training data (all except i)
    train_data <- data[-i, ]
    test_data <- data[i, ]

    # Note: In practice, you'd retrain the model here
    # For demonstration, we'll just use predictions
    pred <- predict_d2h_precip(test_data, model = model_name)
    predictions[i] <- pred$d2h_precip_mean
  }

  # Calculate metrics
  rmse <- sqrt(mean((predictions - data$true_d2h_precip)^2))
  r2 <- cor(predictions, data$true_d2h_precip)^2

  return(list(predictions = predictions, rmse = rmse, r2 = r2))
}
```

### Workflow 3: Publication-Ready Outputs

```{r eval=FALSE}
# Generate publication table
create_publication_table <- function(data, results) {
  pub_table <- data.frame(
    Site = data$site_id,
    Longitude = sprintf("%.2f°%s",
                       abs(data$longitude),
                       ifelse(data$longitude < 0, "W", "E")),
    Latitude = sprintf("%.2f°%s",
                      abs(data$latitude),
                      ifelse(data$latitude < 0, "S", "N")),
    `Elevation (m)` = data$elevation,
    `δ²H_wax (‰)` = sprintf("%.1f ± %.1f",
                            data$d2h_wax,
                            data$d2h_wax_sd),
    `δ²H_precip (‰)` = sprintf("%.1f (%.1f–%.1f)",
                               results$d2h_precip_mean,
                               results$d2h_precip_lower,
                               results$d2h_precip_upper),
    check.names = FALSE
  )

  return(pub_table)
}

# Export to LaTeX
library(xtable)
pub_table <- create_publication_table(example_data[1:5,], results[1:5,])
print(xtable(pub_table), include.rownames = FALSE)
```

## Troubleshooting Performance

### Profiling Code

```{r eval=FALSE}
# Profile predictions
library(profvis)

profvis({
  results <- predict_d2h_precip(large_dataset)
})

# Benchmark different approaches
library(microbenchmark)

microbenchmark(
  direct = predict_d2h_precip(example_data, use_lookup = FALSE),
  lookup = predict_d2h_precip(example_data, use_lookup = TRUE),
  times = 10
)
```

### Optimizing for Production

```{r eval=FALSE}
# Production configuration
setup_production <- function() {
  # Set optimal configuration
  options(
    leafwax.auto_download = FALSE,  # Prevent unexpected downloads
    leafwax.cache_dir = "/data/leafwax_cache",  # Fast SSD
    leafwax.verbose = FALSE  # Reduce output
  )

  # Pre-load frequently used models
  models_to_preload <- c("b0b1", "b0b1_elev", "b0b1_sp")

  for (model in models_to_preload) {
    assign(
      paste0("cached_", model),
      load_posteriors(model, verbose = FALSE),
      envir = .GlobalEnv
    )
  }

  # Create lookup tables
  cache_all_lookup_tables(
    cache_dir = getOption("leafwax.cache_dir"),
    n_draws = 500
  )

  cat("Production environment configured\n")
}

setup_production()
```

## Summary

This vignette covered:

1. **Ensemble methods** for robust predictions
2. **Custom spatial grids** for regional studies
3. **Performance optimization** techniques
4. **Uncertainty quantification** approaches
5. **Integration** with other R packages
6. **Custom workflows** for specific applications

For more information:
- See [Getting Started](getting-started.html) for basics
- See [Model Descriptions](model-descriptions.html) for model details
- Check the [API Reference](../reference/index.html) for function documentation

## Session Info

```{r}
sessionInfo()
```