---
title: "Getting Started with leafwax"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with leafwax}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `leafwax` package provides tools for Bayesian inversion of leaf wax hydrogen isotope measurements (δ²H) to reconstruct past precipitation isotope values. Leaf waxes are long-chain lipid compounds produced by plants that preserve the hydrogen isotopic composition of the water used during their synthesis. By measuring the δ²H of these compounds in sediments or soils, we can reconstruct past hydrological conditions.

This vignette will guide you through:

1. Understanding the available models
2. Performing basic inversions
3. Working with multiple locations
4. Choosing the right model for your data
5. Interpreting uncertainty intervals

## Installation and Setup

```{r setup}
library(leafwax)
```

## Understanding Available Models

The package includes 14 calibration models with different levels of complexity:

```{r models}
# List all available models
models <- available_models()
print(models)

# Get detailed information about a specific model
model_info <- get_model_parameters("baseline_sp")
print(model_info$description)
```

### Model Categories

**Basic Models:**
- `baseline`: Simple OIPC-based calibration without spatial effects
- `baseline_sp`: Basic model with spatial Gaussian process

**Environmental Models:**
- `baseline_env`: Includes elevation effects
- `baseline_env_sp`: Elevation with spatial correlation
- `elevation_only_sp`: Focuses solely on elevation effects

**Vegetation Models:**
- `baseline_veg`: Includes plant functional type (PFT) effects
- `baseline_veg_sp`: PFT with spatial correlation
- `c4_only_sp`: Focuses on C4 vegetation effects

**Complex Models:**
- `full`: Combines elevation and vegetation effects
- `full_sp`: Full model with spatial correlation
- `full_interact`: Includes interaction terms
- `full_interact_sp`: Complete model with all effects

## Basic Example: Single Location

Let's start with a simple inversion at a single location:

```{r single-location}
# Perform inversion with the baseline model
result <- invert_d2h(
  d2h_wax = -150,          # Leaf wax δ²H value (‰)
  d2h_wax_err = 3,         # Measurement uncertainty (‰)
  longitude = -120,        # Longitude (decimal degrees)
  latitude = 40,           # Latitude (decimal degrees)
  model = "baseline"       # Model selection
)

# View the results
print(result)
```

The result contains:
- `d2h_precip_mean`: Mean predicted precipitation δ²H
- `d2h_precip_sd`: Standard deviation of the prediction
- `d2h_precip_lower/upper`: 90% credible interval bounds

## Multiple Locations

The package efficiently handles multiple locations simultaneously:

```{r multi-location}
# Create a dataset with multiple sites
sites <- data.frame(
  site = c("Forest", "Desert", "Mountain", "Coast"),
  d2h_wax = c(-150, -120, -180, -140),
  longitude = c(-120, -115, -105, -124),
  latitude = c(45, 35, 40, 42),
  elevation = c(500, 1200, 2500, 50)
)

# Perform batch inversion
results <- invert_d2h(
  d2h_wax = sites$d2h_wax,
  longitude = sites$longitude,
  latitude = sites$latitude,
  elevation = sites$elevation,
  model = "baseline_env"  # Use model with elevation
)

# Combine with original data
sites_with_results <- cbind(sites, results)
print(sites_with_results[, c("site", "d2h_wax", "d2h_precip_mean", "d2h_precip_sd")])
```

## Choosing the Right Model

The package provides tools to help select the appropriate model based on your available data:

```{r model-selection}
# Get model recommendations based on available data
recommendations <- get_model_recommendations(
  has_elevation = TRUE,    # We have elevation data
  has_c4 = FALSE,          # No C4 vegetation data
  has_pft = FALSE,         # No PFT data
  prefer_spatial = TRUE    # Prefer spatial models for better uncertainty
)

# View top 3 recommendations
for(i in 1:min(3, length(recommendations))) {
  cat("Model", i, ":", names(recommendations)[i], "\n")
  cat("  Score:", recommendations[[i]]$score, "\n")
  cat("  Description:", recommendations[[i]]$description, "\n\n")
}
```

### Validate Your Inputs

Before running an inversion, validate that your data matches the model requirements:

```{r validation}
# Validate inputs for a specific model
validation <- validate_model_inputs(
  model_name = "baseline_env_sp",
  d2h_wax = -150,
  longitude = -120,
  latitude = 40,
  elevation = 1200,
  verbose = TRUE
)
```

## Working with Spatial Models

Spatial models use Gaussian processes with 125 knots on a Fibonacci sphere for better uncertainty quantification:

```{r spatial}
# Compare non-spatial vs spatial model
result_basic <- invert_d2h(
  d2h_wax = -150,
  longitude = -120,
  latitude = 40,
  model = "baseline"
)

result_spatial <- invert_d2h(
  d2h_wax = -150,
  longitude = -120,
  latitude = 40,
  model = "baseline_sp"
)

cat("Basic model:\n")
cat("  Mean:", round(result_basic$d2h_precip_mean, 1), "‰\n")
cat("  SD:", round(result_basic$d2h_precip_sd, 1), "‰\n")
cat("  90% CI width:", round(result_basic$d2h_precip_upper - result_basic$d2h_precip_lower, 1), "‰\n\n")

cat("Spatial model:\n")
cat("  Mean:", round(result_spatial$d2h_precip_mean, 1), "‰\n")
cat("  SD:", round(result_spatial$d2h_precip_sd, 1), "‰\n")
cat("  90% CI width:", round(result_spatial$d2h_precip_upper - result_spatial$d2h_precip_lower, 1), "‰\n")
```

Note that spatial models typically have larger uncertainties because they account for spatial correlation in the residuals.

## Interpreting Uncertainty

The package provides full uncertainty propagation from analytical errors through model parameters:

```{r uncertainty}
# Compare effect of measurement uncertainty
low_error <- invert_d2h(
  d2h_wax = -150,
  d2h_wax_err = 1,  # Low measurement error
  longitude = -120,
  latitude = 40,
  model = "baseline"
)

high_error <- invert_d2h(
  d2h_wax = -150,
  d2h_wax_err = 5,  # High measurement error
  longitude = -120,
  latitude = 40,
  model = "baseline"
)

cat("Effect of measurement uncertainty:\n")
cat("Low error (1‰):\n")
cat("  Prediction SD:", round(low_error$d2h_precip_sd, 1), "‰\n")
cat("High error (5‰):\n")
cat("  Prediction SD:", round(high_error$d2h_precip_sd, 1), "‰\n")
```

## Advanced Usage: Full Posterior Draws

For more detailed analysis, you can access the full posterior distribution:

```{r posterior, eval=FALSE}
# Get full posterior draws (not run in vignette due to size)
result_full <- invert_d2h(
  d2h_wax = -150,
  longitude = -120,
  latitude = 40,
  model = "baseline",
  return_full = TRUE
)

# Access posterior draws
posterior_draws <- result_full$posterior_draws
hist(posterior_draws[1,],
     main = "Posterior Distribution",
     xlab = "Precipitation δ²H (‰)",
     col = "lightblue")
```

## Best Practices

1. **Start simple**: Begin with the `baseline` model to establish a baseline prediction
2. **Add complexity gradually**: Include elevation or vegetation effects if you have the data
3. **Use spatial models** when you need robust uncertainty quantification
4. **Validate inputs**: Always check that your data matches model requirements
5. **Consider uncertainty**: Remember that more complex models may have larger (but more realistic) uncertainties

## Troubleshooting

### Common Issues

**Model not found:**
```{r, eval=FALSE}
# Check available models
models <- available_models()
print(models)
```

**Missing covariates:**
```{r, eval=FALSE}
# The model will warn you about missing required data
# Use a simpler model or provide the missing data
```

**Large uncertainties:**
```{r, eval=FALSE}
# This is often correct - spatial models account for more sources of uncertainty
# Check your measurement errors and consider using non-spatial models if appropriate
```

## Summary

The `leafwax` package provides a comprehensive framework for leaf wax isotope inversion with:

- Multiple models for different data availability scenarios
- Proper uncertainty propagation
- Spatial correlation via Gaussian processes
- Automated model selection tools
- Comprehensive validation functions

For more detailed examples and advanced usage, see the other vignettes and the package documentation.

## References

- Bowen, G. J., et al. (2019). Isotopes in the water cycle. *Annu. Rev. Earth Planet. Sci.* 47, 453-479.
- Sachse, D., et al. (2012). Molecular paleohydrology. *Annu. Rev. Earth Planet. Sci.* 40, 221-249.