---
title: "When does a leaf-wax record support a precipitation-isotope claim?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When does a leaf-wax record support a precipitation-isotope claim?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
set.seed(20260507)
```

The workflow reconstructs `d2H_precip` from a downcore leaf-wax series
and tests whether the reconstructed signal is large enough to support
a published claim of change. The central question for any record is
whether the difference in `d2H_precip` between two stratigraphic
intervals can be distinguished from calibration-plus-analytical noise,
and at what confidence level.

The vignette runs the chain on two Iso2k records that produce opposite
verdicts. Lake Malawi (LS11KOMA, Konecky et al.) has 11 samples across
the Common Era and an in-record `d2H_wax` range of approximately 10
per mil — a record where no plausible 30 per mil shift in `d2H_precip`
can be distinguished from within-record noise. Lake Qinghai
(LS16THQI01, Thomas et al.) has 240 samples spanning 31 kyr of the
glacial-to-Holocene transition on the northeastern Tibetan Plateau,
with strong sample-to-sample autocorrelation that shrinks the
detection threshold; here a substantial drying signal across the
Last Glacial Maximum boundary is recovered with high posterior
probability.

The detection threshold from manuscript Section 4.5.3 is

\[
\mathrm{threshold}_{\mathrm{precip}}
= \frac{1.96 \sqrt{2(1 - \rho_t)}\, \sqrt{\sigma_{\mathrm{residual}}^2 + \sigma_{\mathrm{analytical}}^2}}{\beta_{\mathrm{eff}}}
\]

— the smallest `d2H_precip` change between two independent samples at
this site that can be distinguished from within-record noise at 95
percent confidence. `sigma_residual` (~16 per mil for the spatial
models) is the calibration's posterior residual SD; the spatial GP
intercept contributes equally to every sample in the record and
cancels in any contrast between intervals.

## 1. Load both records

```{r load}
library(leafwax)

malawi_path <- system.file(
  "extdata", "example_records", "LS11KOMA_d2H.csv",
  package = "leafwax"
)
malawi <- read.csv(malawi_path)
malawi$d2h_wax <- malawi$d2H_wax
malawi$age     <- malawi$age_yrBP

qh_path <- system.file(
  "extdata", "example_records", "LS16THQI01_d2H.csv",
  package = "leafwax"
)
qh <- read.csv(qh_path)
qh$d2h_wax <- qh$d2H_wax
qh$age     <- qh$age_yrBP

c(malawi_n        = nrow(malawi),
  malawi_range_pm = round(diff(range(malawi$d2h_wax)), 1),
  qh_n            = nrow(qh),
  qh_range_pm     = round(diff(range(qh$d2h_wax)), 1))
```

Lake Malawi sits at 10.02° S, 34.19° E (477 m); Lake Qinghai at
37.0° N, 100.0° E (3,194 m). The Malawi series covers the Common Era
at roughly 200-yr resolution; the Qinghai series covers the
glacial–Holocene transition at roughly 130-yr resolution.

## 2. Local effective slope

`local_effective_slope()` returns a per-draw vector of the
`d2H_wax`–`d2H_precip` slope at the site, combining the global
posterior `beta_oipc` with the spatial slope GP. The default ceiling
is 0.88, the simple-model upper bound from manuscript Section 4.5.5.
A point-estimate workflow can pass the posterior median; the full
per-draw vector is retained here so slope uncertainty propagates
through the inversion.

```{r slope}
malawi_lon <- 34.1878; malawi_lat <- -10.0183
qh_lon     <- 100;     qh_lat     <- 37

slope_malawi <- suppressWarnings(local_effective_slope(
  longitude = malawi_lon, latitude = malawi_lat,
  model_name = "baseline_sp", n_draws = 100,
  ceiling = 0.88, verbose = FALSE
))

slope_qh <- suppressWarnings(local_effective_slope(
  longitude = qh_lon, latitude = qh_lat,
  model_name = "baseline_sp", n_draws = 100,
  ceiling = 0.88, verbose = FALSE
))

rbind(
  malawi  = quantile(slope_malawi, c(0.025, 0.5, 0.975)),
  qinghai = quantile(slope_qh,     c(0.025, 0.5, 0.975))
)
```

## 3. Bayesian inversion with the local slope

`invert_d2H()` accepts the slope vector from §2. The `record_id`
argument enforces that all rows belong to one site; `return_full = TRUE`
keeps the posterior draws matrix needed by `detect_change()`. Wax-error
sampling uses analytical uncertainty plus the model's posterior
residual SD (supplement S4.1, Eq. 7); the same residual applies to
within-record contrasts because the spatial GP intercept cancels in
any difference between intervals (Section 4.5.3).

```{r invert}
recon_malawi <- suppressWarnings(invert_d2H(
  d2H_wax    = malawi$d2h_wax,
  d2H_wax_sd = rep(3, nrow(malawi)),
  longitude  = rep(malawi_lon, nrow(malawi)),
  latitude   = rep(malawi_lat, nrow(malawi)),
  model_name = "baseline_sp",
  n_posterior_draws = 100,
  slope        = slope_malawi,
  record_id    = "LS11KOMA",
  return_full  = TRUE,
  verbose      = FALSE
))

recon_qh <- suppressWarnings(invert_d2H(
  d2H_wax    = qh$d2h_wax,
  d2H_wax_sd = rep(3, nrow(qh)),
  longitude  = rep(qh_lon, nrow(qh)),
  latitude   = rep(qh_lat, nrow(qh)),
  model_name = "baseline_sp",
  n_posterior_draws = 100,
  slope        = slope_qh,
  record_id    = "LS16THQI01",
  return_full  = TRUE,
  verbose      = FALSE
))
```

## 4. Detection threshold and posterior probability of change

`detect_change()` returns the Section 4.5.3 detection threshold and
the posterior probability that the difference in mean `d2H_precip`
between two intervals exceeds a target magnitude. The threshold uses
the calibration's posterior residual SD (`sigma_residual` ~16 per mil
for spatial models; pull it from `model$draws$sigma` in your fit, or
use the manuscript headline value).

`estimate_temporal_autocorrelation()` returns the lag-1 correlation of
age-ordered residuals after a flat-mean detrend. This is a deliberately
simple AR(1) estimator — accurate on regularly sampled records,
approximate on irregular ones. For irregular paleo series, the
standard tools are `astrochron` (REDFIT and tau-bias correction;
Meyers and colleagues) or Mudelsee's `pearsonT3`, which is
bias-corrected for unevenly spaced data. A Lomb-Scargle estimator is
planned for v0.3 (`method = "lomb_scargle"`).

The Malawi record splits at 1000 yr BP. The Qinghai record splits at
15,000 yr BP — late LGM to mid-Holocene, the boundary across which
many Asian-monsoon records show a regional shift in source-water
`d2H`.

```{r detect}
rho_malawi <- estimate_temporal_autocorrelation(
  malawi$d2h_wax, malawi$age, method = "ar1"
)

dc_malawi <- detect_change(
  reconstruction    = recon_malawi,
  age               = malawi$age,
  baseline_interval = c(0, 1000),
  test_intervals    = list(post_1000 = c(1000, 2000)),
  sigma_residual    = 16,
  sigma_analytical  = 3,
  rho_t             = rho_malawi,
  beta_eff          = stats::median(slope_malawi),
  confidence        = 0.95,
  magnitudes        = c(10, 30, 50)
)

rho_qh <- estimate_temporal_autocorrelation(
  qh$d2h_wax, qh$age, method = "ar1"
)

dc_qh <- detect_change(
  reconstruction    = recon_qh,
  age               = qh$age,
  baseline_interval = c(0, 15000),
  test_intervals    = list(lgm = c(15000, 25000)),
  sigma_residual    = 16,
  sigma_analytical  = 3,
  rho_t             = rho_qh,
  beta_eff          = stats::median(slope_qh),
  confidence        = 0.95,
  magnitudes        = c(10, 30, 50)
)

list(
  malawi  = list(rho_t = round(rho_malawi, 3),
                 threshold_permil = round(dc_malawi$threshold, 1),
                 intervals        = dc_malawi$intervals),
  qinghai = list(rho_t = round(rho_qh, 3),
                 threshold_permil = round(dc_qh$threshold, 1),
                 intervals        = dc_qh$intervals)
)
```

The two records produce opposite verdicts. Malawi: lag-1
autocorrelation `r round(rho_malawi, 2)`, 95 percent detection
threshold approximately `r round(dc_malawi$threshold, 0)` per mil,
posterior probability of a 30 per mil shift across 1000 yr BP
`r sprintf("%.2f", dc_malawi$intervals$p_abs_delta_gt_30)`. Qinghai:
lag-1 autocorrelation `r round(rho_qh, 2)` (densely sampled and
strongly persistent), threshold approximately
`r round(dc_qh$threshold, 0)` per mil, posterior probability of a 30
per mil shift across 15,000 yr BP
`r sprintf("%.2f", dc_qh$intervals$p_abs_delta_gt_30)`. The Malawi
record cannot distinguish a 30 per mil change in `d2H_precip` from
calibration noise; the Qinghai record can. Two factors drive the
contrast: Qinghai's much larger LGM-to-Holocene `d2H_wax` shift, and
its high autocorrelation, which reduces `sqrt(2(1-rho_t))` and pulls
the threshold down.

## 5. Assess a published claim

`assess_claim()` walks the four-level taxonomy from manuscript Section
4.5.6: Level 1 reports the wax-only signal, Level 2 adds the
calibration's detection threshold, Level 3 converts to a magnitude in
`d2H_precip`, Level 4 layers in proxy stationarity and corroborating
evidence. A claim is supported at the highest level where every
prerequisite passes.

```{r assess}
build_claim <- function(beta_eff, rho_t, baseline, test, magnitude_precip) {
  list(
    level             = 4,
    interval_baseline = baseline,
    interval_test     = test,
    sigma_analytical  = 3,
    rho_t             = rho_t,
    confidence        = 0.95,
    beta_eff          = beta_eff,
    magnitude_precip  = magnitude_precip,
    corroborating_proxies = list(
      regional_proxy = "regional records show coeval shift"
    ),
    vegetation_stationary = list(
      value    = TRUE,
      evidence = "n-alkane chain length distributions stable across the boundary"
    ),
    seasonal_source_stationary = list(
      value    = TRUE,
      evidence = "regional d18O record shows no seasonality shift"
    ),
    evapotranspirative_stationary = list(
      value    = TRUE,
      evidence = "leaf-water proxy stable; no aridity transition"
    )
  )
}

malawi_record <- data.frame(
  d2h_wax     = malawi$d2h_wax,
  age         = malawi$age,
  d2h_wax_err = rep(3, nrow(malawi))
)
qh_record <- data.frame(
  d2h_wax     = qh$d2h_wax,
  age         = qh$age,
  d2h_wax_err = rep(3, nrow(qh))
)

verdict_malawi <- suppressWarnings(assess_claim(
  record         = malawi_record,
  claim          = build_claim(stats::median(slope_malawi),
                                rho_malawi,
                                c(0, 1000), c(1000, 2000),
                                magnitude_precip = 30),
  reconstruction = recon_malawi
))

verdict_qh <- suppressWarnings(assess_claim(
  record         = qh_record,
  claim          = build_claim(stats::median(slope_qh),
                                rho_qh,
                                c(0, 15000), c(15000, 25000),
                                magnitude_precip = 30),
  reconstruction = recon_qh
))

c(malawi_highest_level   = verdict_malawi$highest_level,
  malawi_supported_at_4  = verdict_malawi$asserted_supported,
  qinghai_highest_level  = verdict_qh$highest_level,
  qinghai_supported_at_4 = verdict_qh$asserted_supported)
```

Read each `verdict$levels` data frame top to bottom — every level
above the highest one passed is reported with the reason it failed.
The Malawi claim fails at the within-record noise step; the Qinghai
claim clears all four levels because the wax-side LGM-to-Holocene
change is large relative to noise and the asserted stationarity
evidence holds.

## 6. Plot the reconstructions

```{r plot, fig.height = 6, echo = TRUE}
op <- par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

plot_recon <- function(rec, ages, title, boundary) {
  ord <- order(ages)
  plot(
    ages[ord],
    rec$summary$d2h_precip_median[ord],
    type = "o", pch = 16, col = "black",
    xlim = rev(range(ages)),
    ylim = range(c(rec$summary$d2h_precip_lower,
                   rec$summary$d2h_precip_upper)),
    xlab = "Age (yr BP)",
    ylab = expression(delta^2 * H[precipitation] ~ "(‰)"),
    main = title
  )
  polygon(
    c(ages[ord], rev(ages[ord])),
    c(rec$summary$d2h_precip_lower[ord],
      rev(rec$summary$d2h_precip_upper[ord])),
    border = NA, col = adjustcolor("steelblue", alpha.f = 0.3)
  )
  abline(v = boundary, lty = 2, col = "red")
}

plot_recon(recon_malawi, malawi$age,
           "Lake Malawi (LS11KOMA): no detection at 95%",
           boundary = 1000)
plot_recon(recon_qh, qh$age,
           "Lake Qinghai (LS16THQI01): clear LGM signal",
           boundary = 15000)

par(op)
```

Each shaded band is the 90 percent credible interval per sample,
propagating analytical uncertainty, regression-parameter uncertainty,
the local slope posterior, and the calibration's posterior residual
SD. The dashed red line marks the interval boundary used in §4 and §5.

## Takeaway

A record's ability to support a `d2H_precip`-change claim depends on
three quantities computable from the record alone: its within-record
`d2H_wax` interval-mean contrast relative to `sigma_residual`, the
local effective slope, and the lag-1 temporal autocorrelation.
`detect_change()` packages these into a single threshold and a
posterior probability; `assess_claim()` walks them through the
four-level taxonomy. Records like Malawi do not clear Level 2 for
plausible 30 per mil claims; long, densely-sampled records like
Qinghai do, because their large interval-mean shifts and high
sample-to-sample autocorrelation jointly raise signal-to-noise above
the calibration's noise floor.

## Notes

- The vignette uses 100 posterior draws for speed. Production
  reconstructions should use the full posterior (omit `n_draws` and
  `n_posterior_draws`).
- For irregular paleo series, replace the built-in
  `estimate_temporal_autocorrelation()` with `astrochron::redfit` or
  the bias-corrected `pearsonT3` from Mudelsee (2002). The leafwax
  built-in is a flat-mean lag-1 estimator and assumes near-regular
  spacing.
