---
title: "Paleo-record workflow: from a downcore series to a defensible claim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Paleo-record workflow: from a downcore series to a defensible claim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
set.seed(20260507)
```

This vignette walks the paleo-record workflow on a real Iso2k
record, Lake Malawi (LS11KOMA, Konecky et al.). The chain is:

1. Load the downcore series.
2. Extract or override the local effective slope.
3. Run the inversion with `record_id` and `slope`.
4. Compute a within-record detection threshold and the posterior
   probability of a hypothesised change.
5. Assess a published claim against the four-level taxonomy from
   manuscript Section 4.5.6.
6. Plot the reconstructed `d2H_precip` series.

The detection threshold comes from the calibration's posterior
residual SD (`sigma_residual`, approximately 16 per mil for the
spatial models), combined with analytical uncertainty and the local
effective slope (manuscript Section 4.5.3). The spatial GP
intercept's contribution cancels in any contrast between intervals
of the same record.

## 1. Load the downcore series

```{r load}
library(leafwax)

malawi_path <- system.file(
  "extdata", "example_records", "LS11KOMA_d2H.csv",
  package = "leafwax"
)
malawi <- read.csv(malawi_path)
malawi$d2h_wax <- malawi$d2H_wax
malawi$age     <- malawi$age_yrBP
malawi
```

The record has 11 samples spanning the Common Era (~100--2000 yrBP) at
Lake Malawi (10.02 deg S, 34.19 deg E).

## 2. Local effective slope

`local_effective_slope()` returns a per-draw vector of the
d2H_wax-d2H_precip slope at the site, combining the global posterior
beta_oipc with the spatial slope GP. The default ceiling is 0.88,
the simple-model upper bound from manuscript Section 4.5.5.

```{r slope}
malawi_lon <- 34.1878
malawi_lat <- -10.0183

slope_post <- suppressWarnings(local_effective_slope(
  longitude   = malawi_lon,
  latitude    = malawi_lat,
  model_name  = "baseline_sp",
  n_draws     = 100,
  ceiling     = 0.88,
  verbose     = FALSE
))

quantile(slope_post, c(0.025, 0.5, 0.975))
```

A defended slope point estimate would typically be the posterior
median; here we keep the per-draw vector so the inversion propagates
slope uncertainty.

## 3. Bayesian inversion with the local slope

`invert_d2H()` accepts the slope vector built above. We pass
`record_id` so the function validates that all rows are from one
site, and we ask for the full posterior draws matrix so we can hand
it to `detect_change()`. The wax-error draw uses analytical
uncertainty plus the model's posterior residual SD (manuscript
supplement Section S4.1, Eq. 7); the same residual applies to
within-record contrasts because the spatial GP intercept's
contribution cancels in any difference between time intervals
(manuscript Section 4.5.3).

```{r invert}
recon <- suppressWarnings(invert_d2H(
  d2H_wax    = malawi$d2h_wax,
  d2H_wax_sd = rep(3, nrow(malawi)),
  longitude  = rep(malawi_lon, nrow(malawi)),
  latitude   = rep(malawi_lat, nrow(malawi)),
  model_name = "baseline_sp",
  n_posterior_draws = 100,
  slope        = slope_post,
  record_id    = "LS11KOMA",
  return_full  = TRUE,
  verbose      = FALSE
))

head(recon$summary[, c("d2h_wax", "d2h_precip_median",
                       "d2h_precip_lower", "d2h_precip_upper")])
```

The returned `summary` has one row per sample with the posterior
median and credible-interval bounds; `posterior_draws` is the
n_iter x n_samples matrix used below.

## 4. Detection threshold and posterior probability of change

`detect_change()` reports the manuscript Section 4.5.3 detection
threshold and the posterior probability that the difference in mean
d2H_precip between two intervals exceeds a magnitude. We split the
record at 1000 yrBP and ask whether there was a 30 per mil shift in
d2H_precip across the boundary. The threshold uses the calibration's
posterior residual SD (`sigma_residual` ~16 per mil for spatial
models; pull it from `model$draws$sigma` in your fit, or use the
manuscript headline value).

```{r detect}
rho_t <- estimate_temporal_autocorrelation(
  malawi$d2h_wax, malawi$age, method = "ar1"
)

dc <- detect_change(
  reconstruction    = recon,
  age               = malawi$age,
  baseline_interval = c(0, 1000),
  test_intervals    = list(post_1000 = c(1000, 2000)),
  sigma_residual    = 16,
  sigma_analytical  = 3,
  rho_t             = rho_t,
  beta_eff          = stats::median(slope_post),
  confidence        = 0.95,
  magnitudes        = c(10, 30, 50)
)

dc$threshold
dc$intervals
```

The threshold is the smallest d2H_precip change between two
independent samples at this site that can be distinguished from
calibration noise at 95% confidence. The record's actual posterior
delta is typically smaller, which is the point of the example.

## 5. Assess a published claim

`assess_claim()` walks the four-level taxonomy from manuscript
Section 4.5.6. We test a hypothetical claim that this record shows
a 30 per mil drying signal across 1000 yrBP, supported by some
corroborating evidence and stationarity arguments.

```{r assess}
malawi_record <- data.frame(
  d2h_wax     = malawi$d2h_wax,
  age         = malawi$age,
  d2h_wax_err = rep(3, nrow(malawi))
)

claim <- list(
  level             = 4,                       # asserted level
  interval_baseline = c(0, 1000),
  interval_test     = c(1000, 2000),
  sigma_analytical  = 3,
  rho_t             = rho_t,
  confidence        = 0.95,
  beta_eff          = stats::median(slope_post),
  magnitude_precip  = 30,
  corroborating_proxies = list(
    speleothem_d18O = "regional speleothems show coeval shift"
  ),
  vegetation_stationary = list(
    value    = TRUE,
    evidence = "n-alkane chain length distributions stable across the boundary"
  ),
  seasonal_source_stationary = list(
    value    = TRUE,
    evidence = "regional speleothem d18O shows no seasonality shift"
  ),
  evapotranspirative_stationary = list(
    value    = TRUE,
    evidence = "leaf-water proxy stable; no aridity transition"
  )
)

verdict <- suppressWarnings(assess_claim(
  record         = malawi_record,
  claim          = claim,
  reconstruction = recon
))

verdict$highest_level
verdict$asserted_supported
verdict$levels
```

The taxonomy walks each level in turn: every level above the highest
one passed is reported with the reason it failed. Reading the
`levels` data frame top to bottom is the recommended way to see what
a record-claim pair supports.

## 6. Plot the reconstructed d2H_precip

```{r plot, echo = TRUE}
op <- par(mar = c(4, 4, 1, 1))

ord <- order(malawi$age)
plot(
  malawi$age[ord],
  recon$summary$d2h_precip_median[ord],
  type = "o", pch = 16, col = "black",
  xlim = rev(range(malawi$age)),
  ylim = range(c(recon$summary$d2h_precip_lower,
                 recon$summary$d2h_precip_upper)),
  xlab = "Age (yr BP)",
  ylab = "d2H_precip (per mil)"
)
polygon(
  c(malawi$age[ord], rev(malawi$age[ord])),
  c(recon$summary$d2h_precip_lower[ord],
    rev(recon$summary$d2h_precip_upper[ord])),
  border = NA, col = adjustcolor("steelblue", alpha.f = 0.3)
)
par(op)
```

The shaded band is the 90% credible interval per sample, propagating
analytical uncertainty, regression-parameter uncertainty, the local
slope posterior, and the calibration's posterior residual SD.

## Notes

- This vignette uses 200 posterior draws for speed. Production
  reconstructions should use the full posterior (omit `n_draws`
  and `n_posterior_draws`).
- Re-running the workflow on a longer record (e.g., LS16THN301, an
  8 kyr Holocene series with larger d2H_precip changes) reaches
  Level 3+ for the larger claimed shifts that record was published
  with. The workflow is identical; only the inputs change.
