---
title: "When does a leaf-wax record support a precipitation-isotope claim?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When does a leaf-wax record support a precipitation-isotope claim?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
set.seed(20260507)
```

The workflow reconstructs `d2H_precip` from a downcore leaf-wax series
and tests whether the reconstructed signal is large enough to support
a published claim of change. The central question for any record is
whether the difference in `d2H_precip` between two stratigraphic
intervals can be distinguished from calibration-plus-analytical noise,
and at what confidence level.

The vignette runs the chain on two Iso2k records with different
inferential ceilings. Zaca Lake (LS14FEZA, Heusser et al.) has 518
samples spanning the last 3 kyr in coastal southern California and an
in-record `d2H_wax` range of approximately 77 per mil distributed
without a coherent secular trend — a record where no plausible 30 per
mil shift in `d2H_precip` can be distinguished from within-record
noise across any reasonable interval split. Lake Qinghai (LS16THQI01,
Thomas et al.) has 240 samples spanning 31 kyr of the
glacial-to-Holocene transition on the northeastern Tibetan Plateau.
Its large LGM-to-Holocene shift and strong sample-to-sample
autocorrelation support a directional hydroclimate-change claim, but
the example 30 per mil quantitative `d2H_precip` claim still does not
clear the 95 percent Level 3 threshold.

The detection threshold from manuscript Section 4.5.3 is

\[
\mathrm{threshold}_{\mathrm{precip}}
= \frac{1.96 \sqrt{2(1 - \rho_t)}\, \sqrt{\sigma_{\mathrm{residual}}^2 + \sigma_{\mathrm{analytical}}^2}}{\beta_{\mathrm{eff}}}
\]

— the smallest `d2H_precip` change between two independent samples at
this site that can be distinguished from within-record noise at 95
percent confidence. `sigma_residual` (~16 per mil for the spatial
models) is the calibration's posterior residual SD; the spatial GP
intercept contributes equally to every sample in the record and
cancels in any contrast between intervals.

## 1. Load both records

```{r load}
library(leafwax)

record_path <- function(filename) {
  path <- system.file(
    "extdata", "example_records", filename,
    package = "leafwax"
  )
  if (nzchar(path)) return(path)

  local_path <- file.path(
    "..", "inst", "extdata", "example_records", filename
  )
  if (file.exists(local_path)) return(local_path)

  stop("Could not locate example record: ", filename)
}

zaca_path <- record_path("LS14FEZA_d2H.csv")
zaca <- read.csv(zaca_path)
zaca$d2h_wax <- zaca$d2H_wax
zaca$age     <- zaca$age_yrBP

qh_path <- record_path("LS16THQI01_d2H.csv")
qh <- read.csv(qh_path)
qh$d2h_wax <- qh$d2H_wax
qh$age     <- qh$age_yrBP

c(zaca_n        = nrow(zaca),
  zaca_range_pm = round(diff(range(zaca$d2h_wax)), 1),
  qh_n            = nrow(qh),
  qh_range_pm     = round(diff(range(qh$d2h_wax)), 1))
```

Zaca Lake sits at 34.78° N, 120.04° W (730 m) on the southern
California coast; Lake Qinghai at 37.0° N, 100.0° E (3,194 m) on the
northeastern Tibetan Plateau. The Zaca series covers the last 3 kyr
at roughly 6-yr resolution; the Qinghai series covers the
glacial–Holocene transition at roughly 130-yr resolution.

## 2. Local effective slope

`local_effective_slope()` returns a per-draw vector of the
`d2H_wax`–`d2H_precip` slope at the site, combining the global
posterior `beta_oipc` with the spatial slope GP. The default ceiling
is 0.88, the simple-model upper bound from manuscript Section 4.5.5.
A point-estimate workflow can pass the posterior median; the full
per-draw vector is retained here so slope uncertainty propagates
through the inversion.

```{r slope}
zaca_lon <- -120.0392; zaca_lat <- 34.7778
qh_lon     <- 100;     qh_lat     <- 37

slope_zaca <- suppressWarnings(local_effective_slope(
  longitude = zaca_lon, latitude = zaca_lat,
  model_name = "baseline_sp", n_draws = 100,
  ceiling = 0.88, verbose = FALSE
))

slope_qh <- suppressWarnings(local_effective_slope(
  longitude = qh_lon, latitude = qh_lat,
  model_name = "baseline_sp", n_draws = 100,
  ceiling = 0.88, verbose = FALSE
))

rbind(
  zaca  = quantile(slope_zaca, c(0.025, 0.5, 0.975)),
  qinghai = quantile(slope_qh,     c(0.025, 0.5, 0.975))
)
```

## 3. Bayesian inversion with the local slope

`invert_d2H()` accepts the slope vector from §2. The `record_id`
argument enforces that all rows belong to one site; `return_full = TRUE`
keeps the posterior draws matrix needed by `detect_change()`. Wax-error
sampling uses analytical uncertainty plus the model's posterior
residual SD (supplement S4.1, Eq. 7); the same residual applies to
within-record contrasts because the spatial GP intercept cancels in
any difference between intervals (Section 4.5.3).

```{r invert}
recon_zaca <- suppressWarnings(invert_d2H(
  d2H_wax    = zaca$d2h_wax,
  d2H_wax_sd = rep(3, nrow(zaca)),
  longitude  = rep(zaca_lon, nrow(zaca)),
  latitude   = rep(zaca_lat, nrow(zaca)),
  model_name = "baseline_sp",
  n_posterior_draws = 100,
  slope        = slope_zaca,
  record_id    = "LS14FEZA",
  return_full  = TRUE,
  verbose      = FALSE
))

recon_qh <- suppressWarnings(invert_d2H(
  d2H_wax    = qh$d2h_wax,
  d2H_wax_sd = rep(3, nrow(qh)),
  longitude  = rep(qh_lon, nrow(qh)),
  latitude   = rep(qh_lat, nrow(qh)),
  model_name = "baseline_sp",
  n_posterior_draws = 100,
  slope        = slope_qh,
  record_id    = "LS16THQI01",
  return_full  = TRUE,
  verbose      = FALSE
))
```

## 4. Detection threshold and posterior probability of change

`detect_change()` returns the Section 4.5.3 detection threshold and
the posterior probability that the difference in mean `d2H_precip`
between two intervals exceeds a target magnitude. The threshold uses
the calibration's posterior residual SD (`sigma_residual` ~16 per mil
for spatial models; pull it from `model$draws$sigma` in your fit, or
use the manuscript headline value).

`estimate_temporal_autocorrelation()` returns the lag-1 correlation of
age-ordered residuals after a flat-mean detrend. This is a deliberately
simple AR(1) estimator — accurate on regularly sampled records,
approximate on irregular ones. For irregular paleo series, the
standard tools are `astrochron` (REDFIT and tau-bias correction;
Meyers and colleagues) or Mudelsee's `pearsonT3`, which is
bias-corrected for unevenly spaced data. A Lomb-Scargle estimator is
planned for v0.3 (`method = "lomb_scargle"`).

The Zaca record splits at 1000 yr BP. The Qinghai record splits at
15,000 yr BP — late LGM to mid-Holocene, the boundary across which
many Asian-monsoon records show a regional shift in source-water
`d2H`.

```{r detect}
rho_zaca <- estimate_temporal_autocorrelation(
  zaca$d2h_wax, zaca$age, method = "ar1"
)

dc_zaca <- detect_change(
  reconstruction    = recon_zaca,
  age               = zaca$age,
  baseline_interval = c(0, 1000),
  test_intervals    = list(post_1000 = c(1000, 2000)),
  sigma_residual    = 16,
  sigma_analytical  = 3,
  rho_t             = rho_zaca,
  beta_eff          = stats::median(slope_zaca),
  confidence        = 0.95,
  magnitudes        = c(10, 30, 50)
)

rho_qh <- estimate_temporal_autocorrelation(
  qh$d2h_wax, qh$age, method = "ar1"
)

dc_qh <- detect_change(
  reconstruction    = recon_qh,
  age               = qh$age,
  baseline_interval = c(0, 15000),
  test_intervals    = list(lgm = c(15000, 25000)),
  sigma_residual    = 16,
  sigma_analytical  = 3,
  rho_t             = rho_qh,
  beta_eff          = stats::median(slope_qh),
  confidence        = 0.95,
  magnitudes        = c(10, 30, 50)
)

list(
  zaca  = list(rho_t = round(rho_zaca, 3),
                 threshold_permil = round(dc_zaca$threshold, 1),
                 intervals        = dc_zaca$intervals),
  qinghai = list(rho_t = round(rho_qh, 3),
                 threshold_permil = round(dc_qh$threshold, 1),
                 intervals        = dc_qh$intervals)
)
```

The two records produce different verdicts. Zaca: lag-1
autocorrelation `r round(rho_zaca, 2)`, 95 percent detection
threshold approximately `r round(dc_zaca$threshold, 0)` per mil,
posterior probability of a 30 per mil shift across 1000 yr BP
`r sprintf("%.2f", dc_zaca$intervals$p_abs_delta_gt_30)`. Qinghai:
lag-1 autocorrelation `r round(rho_qh, 2)` (densely sampled and
strongly persistent), threshold approximately
`r round(dc_qh$threshold, 0)` per mil, posterior probability of a 30
per mil shift across 15,000 yr BP
`r sprintf("%.2f", dc_qh$intervals$p_abs_delta_gt_30)`. The Zaca
record cannot distinguish a 30 per mil change in `d2H_precip` from
calibration noise. The Qinghai record provides much stronger evidence
for a large LGM-to-Holocene shift, but the 30 per mil quantitative
claim remains below the 95 percent decision threshold. Two factors
drive the contrast: Qinghai's much larger LGM-to-Holocene `d2H_wax`
shift, and its high autocorrelation, which reduces `sqrt(2(1-rho_t))`
and pulls the threshold down.

## 5. Assess a published claim

`assess_claim()` walks the four-level taxonomy from manuscript Section
4.5.6: Level 1 reports the wax-only signal, Level 2 adds the
calibration's detection threshold, Level 3 converts to a magnitude in
`d2H_precip`, Level 4 layers in proxy stationarity and corroborating
evidence. A claim is supported at the highest level where every
prerequisite passes.

```{r assess}
build_claim <- function(beta_eff, rho_t, baseline, test, magnitude_precip) {
  list(
    level             = 4,
    interval_baseline = baseline,
    interval_test     = test,
    sigma_analytical  = 3,
    rho_t             = rho_t,
    confidence        = 0.95,
    beta_eff          = beta_eff,
    magnitude_precip  = magnitude_precip,
    corroborating_proxies = list(
      regional_proxy = "regional records show coeval shift"
    ),
    vegetation_stationary = list(
      value    = TRUE,
      evidence = "n-alkane chain length distributions stable across the boundary"
    ),
    seasonal_source_stationary = list(
      value    = TRUE,
      evidence = "regional d18O record shows no seasonality shift"
    ),
    evapotranspirative_stationary = list(
      value    = TRUE,
      evidence = "leaf-water proxy stable; no aridity transition"
    )
  )
}

zaca_record <- data.frame(
  d2h_wax     = zaca$d2h_wax,
  age         = zaca$age,
  d2h_wax_err = rep(3, nrow(zaca))
)
qh_record <- data.frame(
  d2h_wax     = qh$d2h_wax,
  age         = qh$age,
  d2h_wax_err = rep(3, nrow(qh))
)

verdict_zaca <- suppressWarnings(assess_claim(
  record         = zaca_record,
  claim          = build_claim(stats::median(slope_zaca),
                                rho_zaca,
                                c(0, 1000), c(1000, 2000),
                                magnitude_precip = 30),
  reconstruction = recon_zaca
))

verdict_qh <- suppressWarnings(assess_claim(
  record         = qh_record,
  claim          = build_claim(stats::median(slope_qh),
                                rho_qh,
                                c(0, 15000), c(15000, 25000),
                                magnitude_precip = 30),
  reconstruction = recon_qh
))

c(zaca_highest_level   = verdict_zaca$highest_level,
  zaca_supported_at_4  = verdict_zaca$asserted_supported,
  qinghai_highest_level  = verdict_qh$highest_level,
  qinghai_supported_at_4 = verdict_qh$asserted_supported)
```

Read each `verdict$levels` data frame top to bottom — every level
above the highest one passed is reported with the reason it failed.
The Zaca claim fails at the within-record noise step; the Qinghai
claim clears Level 2 for the asserted 30 per mil Level 4 claim. It
fails Level 3 because the posterior probability for a 30 per mil
`d2H_precip` shift is below 0.95, even though the example supplies
stationarity evidence for the Level 4 controls.

## 6. Plot the reconstructions

```{r plot, fig.height = 6, echo = TRUE}
op <- par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

plot_recon <- function(rec, ages, title, boundary) {
  ord <- order(ages)
  plot(
    ages[ord],
    rec$summary$d2h_precip_median[ord],
    type = "o", pch = 16, col = "black",
    xlim = rev(range(ages)),
    ylim = range(c(rec$summary$d2h_precip_lower,
                   rec$summary$d2h_precip_upper)),
    xlab = "Age (yr BP)",
    ylab = expression(delta^2 * H[precipitation] ~ "(‰)"),
    main = title
  )
  polygon(
    c(ages[ord], rev(ages[ord])),
    c(rec$summary$d2h_precip_lower[ord],
      rev(rec$summary$d2h_precip_upper[ord])),
    border = NA, col = adjustcolor("steelblue", alpha.f = 0.3)
  )
  abline(v = boundary, lty = 2, col = "red")
}

plot_recon(recon_zaca, zaca$age,
           "Zaca Lake (LS14FEZA): no detection at 95%",
           boundary = 1000)
plot_recon(recon_qh, qh$age,
           "Lake Qinghai (LS16THQI01): large LGM shift",
           boundary = 15000)

par(op)
```

Each shaded band is the 90 percent credible interval per sample,
propagating analytical uncertainty, regression-parameter uncertainty,
the local slope posterior, and the calibration's posterior residual
SD. The dashed red line marks the interval boundary used in §4 and §5.

## Takeaway

A record's ability to support a `d2H_precip`-change claim depends on
three quantities computable from the record alone: its within-record
`d2H_wax` interval-mean contrast relative to `sigma_residual`, the
local effective slope, and the lag-1 temporal autocorrelation.
`detect_change()` packages these into a single threshold and a
posterior probability; `assess_claim()` walks them through the
four-level taxonomy. Records like Zaca do not clear Level 2 for
plausible 30 per mil claims. Long, densely sampled records like
Qinghai can clear directional hydroclimate-change claims and provide
much stronger quantitative evidence, but the example 30 per mil Level
4 claim still requires posterior support at the chosen confidence
level.

## Notes

- The vignette uses 100 posterior draws for speed. Production
  reconstructions should use the full posterior (omit `n_draws` and
  `n_posterior_draws`).
- For irregular paleo series, replace the built-in
  `estimate_temporal_autocorrelation()` with `astrochron::redfit` or
  the bias-corrected `pearsonT3` from Mudelsee (2002). The leafwax
  built-in is a flat-mean lag-1 estimator and assumes near-regular
  spacing.
