---
title: "Paleo-record workflow: from a downcore series to a defensible claim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Paleo-record workflow: from a downcore series to a defensible claim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
set.seed(20260507)
```

This vignette walks the v0.2.0 paleo-record workflow on a real Iso2k
record, Lake Malawi (LS11KOMA, Konecky et al.). The chain is:

1. Load the downcore series.
2. Estimate the within-record residual SD (`sigma_within`).
3. Extract or override the local effective slope.
4. Run the inversion with `record_id`, `sigma_within`, and `slope`.
5. Compute a within-record detection threshold and the posterior
   probability of a hypothesised change.
6. Assess a published claim against the four-level taxonomy from
   manuscript Section 4.5.6.
7. Plot the reconstructed `d2H_precip` series with within-record
   uncertainty.

Lake Malawi was chosen because its within-record variability is
small (~3 per mil over the Common Era), which makes it a useful
educational example: most claimed shifts at this site will not
survive Level 1.

## 1. Load the downcore series

```{r load}
library(leafwax)

malawi_path <- system.file(
  "extdata", "example_records", "LS11KOMA_d2H.csv",
  package = "leafwax"
)
malawi <- read.csv(malawi_path)
malawi$d2h_wax <- malawi$d2H_wax
malawi$age     <- malawi$age_yrBP
malawi
```

The record has 11 samples spanning the Common Era (~100--2000 yrBP) at
Lake Malawi (10.02 deg S, 34.19 deg E).

## 2. Estimate sigma_within

`estimate_sigma_within()` computes the within-record residual SD on a
stationarity-defended baseline interval. Manuscript Section 4.5.3
explains why the global posterior `sigma` (~16 per mil) overstates the
relevant noise for change detection inside a single record.

For Lake Malawi, the full Common-Era record is reasonably stationary
and we treat the entire 11-sample series as the baseline. The
function will warn that this conflates real climate variability with
noise -- accepted here for illustration; a real application should
defend a sub-interval on independent grounds.

```{r sigma_within}
sw <- suppressWarnings(estimate_sigma_within(
  d2h_wax       = malawi$d2h_wax,
  age           = malawi$age,
  baseline_interval = NULL,   # full record (warning)
  detrend       = "linear",
  ar1_correction = TRUE
))
sw
```

A within-record SD of ~3 per mil after linear detrending is
characteristic of Common-Era leaf-wax records.

## 3. Local effective slope

`local_effective_slope()` returns a per-draw vector of the
d2H_wax-d2H_precip slope at the site, combining the global posterior
beta_oipc with the spatial slope GP. The default ceiling is 0.88,
the simple-model upper bound from manuscript Section 4.5.5.

```{r slope}
malawi_lon <- 34.1878
malawi_lat <- -10.0183

slope_post <- suppressWarnings(local_effective_slope(
  longitude   = malawi_lon,
  latitude    = malawi_lat,
  model_name  = "baseline_sp",
  n_draws     = 200,
  ceiling     = 0.88,
  verbose     = FALSE
))

quantile(slope_post, c(0.025, 0.5, 0.975))
```

A defended slope point estimate would typically be the posterior
median; here we keep the per-draw vector so the inversion propagates
slope uncertainty.

## 4. Bayesian inversion with sigma_within and slope

`invert_d2H()` accepts the slope vector built above plus the within-
record SD. We pass `record_id` so the function validates that all
rows are from one site, and we ask for the full posterior draws
matrix so we can hand it to `detect_change()`.

```{r invert}
recon <- suppressWarnings(invert_d2H(
  d2H_wax    = malawi$d2h_wax,
  d2H_wax_sd = rep(3, nrow(malawi)),
  longitude  = rep(malawi_lon, nrow(malawi)),
  latitude   = rep(malawi_lat, nrow(malawi)),
  model_name = "baseline_sp",
  n_posterior_draws = 200,
  sigma_within = sw$sigma_within,
  slope        = slope_post,
  record_id    = "LS11KOMA",
  return_full  = TRUE,
  verbose      = FALSE
))

head(recon$summary[, c("d2h_wax", "d2h_precip_median",
                       "d2h_precip_lower", "d2h_precip_upper")])
```

The returned `summary` has one row per sample with the posterior
median and credible-interval bounds; `posterior_draws` is the
n_iter x n_samples matrix used below.

## 5. Detection threshold and posterior probability of change

`detect_change()` reports the manuscript Section 4.5.3 detection
threshold and the posterior probability that the difference in mean
d2H_precip between two intervals exceeds a magnitude. We split the
record at 1000 yrBP and ask whether there was a 30 per mil shift in
d2H_precip across the boundary.

```{r detect}
rho_t <- estimate_temporal_autocorrelation(
  malawi$d2h_wax, malawi$age, method = "ar1"
)

dc <- detect_change(
  reconstruction    = recon,
  age               = malawi$age,
  baseline_interval = c(0, 1000),
  test_intervals    = list(post_1000 = c(1000, 2000)),
  sigma_within      = sw$sigma_within,
  sigma_analytical  = 3,
  rho_t             = rho_t,
  beta_eff          = stats::median(slope_post),
  confidence        = 0.95,
  magnitudes        = c(10, 30, 50)
)

dc$threshold
dc$intervals
```

The threshold is the smallest d2H_precip change between two
independent samples at this site that can be distinguished from
within-record noise at 95% confidence. With sigma_within ~3 per mil
the threshold is comparatively small; the actual record's posterior
delta is typically smaller still, which is the point of the example.

## 6. Assess a published claim

`assess_claim()` walks the four-level taxonomy from manuscript
Section 4.5.6. We test a hypothetical claim that this record shows
a 30 per mil drying signal across 1000 yrBP, supported by some
corroborating evidence and stationarity arguments.

```{r assess}
malawi_record <- data.frame(
  d2h_wax     = malawi$d2h_wax,
  age         = malawi$age,
  d2h_wax_err = rep(3, nrow(malawi))
)

claim <- list(
  level             = 4,                       # asserted level
  interval_baseline = c(0, 1000),
  interval_test     = c(1000, 2000),
  sigma_within      = sw$sigma_within,
  sigma_analytical  = 3,
  rho_t             = rho_t,
  confidence        = 0.95,
  beta_eff          = stats::median(slope_post),
  magnitude_precip  = 30,
  corroborating_proxies = list(
    speleothem_d18O = "regional speleothems show coeval shift"
  ),
  vegetation_stationary = list(
    value    = TRUE,
    evidence = "n-alkane chain length distributions stable across the boundary"
  ),
  seasonal_source_stationary = list(
    value    = TRUE,
    evidence = "regional speleothem d18O shows no seasonality shift"
  ),
  evapotranspirative_stationary = list(
    value    = TRUE,
    evidence = "leaf-water proxy stable; no aridity transition"
  )
)

verdict <- suppressWarnings(assess_claim(
  record         = malawi_record,
  claim          = claim,
  reconstruction = recon
))

verdict$highest_level
verdict$asserted_supported
verdict$levels
```

The taxonomy walks each level in turn: every level above the highest
one passed is reported with the reason it failed. Reading the
`levels` data frame top to bottom is the recommended way to see what
a record-claim pair supports.

## 7. Plot the reconstructed d2H_precip with within-record uncertainty

```{r plot, echo = TRUE}
op <- par(mar = c(4, 4, 1, 1))

ord <- order(malawi$age)
plot(
  malawi$age[ord],
  recon$summary$d2h_precip_median[ord],
  type = "o", pch = 16, col = "black",
  xlim = rev(range(malawi$age)),
  ylim = range(c(recon$summary$d2h_precip_lower,
                 recon$summary$d2h_precip_upper)),
  xlab = "Age (yr BP)",
  ylab = "d2H_precip (per mil)"
)
polygon(
  c(malawi$age[ord], rev(malawi$age[ord])),
  c(recon$summary$d2h_precip_lower[ord],
    rev(recon$summary$d2h_precip_upper[ord])),
  border = NA, col = adjustcolor("steelblue", alpha.f = 0.3)
)
par(op)
```

The shaded band is the 90% credible interval per sample, propagating
analytical uncertainty, regression-parameter uncertainty, the local
slope posterior, and the within-record residual SD.

## Notes

- This vignette uses 200 posterior draws for speed. Production
  reconstructions should use the full posterior (omit `n_draws`
  and `n_posterior_draws`).
- Re-running the workflow on a longer record (e.g., LS16THN301, an
  8 kyr Holocene series with ~35 per mil within-record SD) returns a
  much larger detection threshold and reaches Level 3+ for the
  larger claimed shifts that record was published with. The
  workflow is identical; only the inputs change.
